---
title: "Reproduction of a Validation Study for a Group-Based Assessment of Children's Executive Functions (2017, Journal of Applied Developmental Psychology)"
author: "Lily Steyer (lsteyer@stanford.edu)"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: false
---

## Introduction

### Justification

This study validated a scalable group-based procedure for assessing elementary school students' executive functions in classroom settings. It was authored by my advisor in the Graduate School of Education, Jelena Obradovic, alongside our lab's current research scientist and several former doctoral students. With Jelena's input, I have chosen to reproduce this study's analysis for several key reasons: (1) it will give me direct experience with the data analysis methods commonly used in our lab, enabling me to contribute to similar future analyses; (2) it is an appropriate level of challenge for me, given that I have no prior experience coding in R and a very basic statistical background relative to my lab colleagues; (3) it will introduce me to key constructs in children's executive functions (EF), as well as the the relationships between those constructs; and (4) it will prepare me to contribute to several current projects in the lab wherein we are using a version of this very assessment protocol to measure children's longitudinal EF development. The members of my lab felt that this reproduction would be more relevant to our work than a separate replication project. 

### Data and Procedures

I will be using the large original dataset from this validation study, which is currently privately housed within my lab but which I hope to help make publically available online as a part of my reproduction. I will adhere to the complex four-step analysis plan used in the original paper, which included:

* Examining descriptive statistics and the correlations between individual assessment (IA) and group assessment (GA) for each EF task
* Comparing associations of EF composites with known demographic and educational covariates across IA and GA settings
* Determining whether there were mean differences between IA and GA performance composites while accounting for the ordering of the two assessments  
* Testing the unique contribution of EF performance in IA and GA settings for the prediction of self- regulated classroom behaviors and academic achievement

### Anticipated Challenges

The largest anticipated challenge involves simultaneously familiarizing myself with R, which I have never coded in before, and with the original paper's statistical methods, some of which are significantly more complex than methods I have used in the past. 

### Relevant Links

**Project Repository:** [https://github.com/psych251/obradovic2017](https://github.com/psych251/obradovic2017)

**Original Paper:** [https://github.com/psych251/obradovic2017/tree/master/original_paper](https://github.com/psych251/obradovic2017/tree/master/original_paper) 

## Methods

### Power Analysis

*To be filled in: Original effect size, power analysis for samples to achieve 80%, 90%, 95% power to detect that effect size. Considerations of feasibility for selecting planned sample size.*

### Planned Sample

*To be filled in: Planned sample size and/or termination rule, sampling frame, known demographics if any, preselection rules if any.*

### Materials 

The measures described below (quoted and followed precisely from the original article) were used to collect this data set: 

#### Hearts and Flowers

"The Hearts and Flowers (H&F) task, designed to assess inhibitory control and cognitive flexibility skills, has been widely used with elementary school students (Davidson, Amso, Anderson, & Diamond, 2006; Oberle & Schonert-Reichl, 2013; Roy, McCoy, & Raver, 2014; Yeniad et al., 2014). There were three blocks: (1) 12 congruent ‘heart’ trials, (2) 12 incongruent ‘flower’ trials, and (3) 33 mixed ‘heart and flower’ trials. Students were presented with an image of a red heart or flower on one side of the screen. For congruent heart trials, students were instructed to press the button on the same side as the presented stimuli (i.e., heart). For incongruent flower trials, students were instructed to press the button on the opposite side of the stimuli (i.e., flower). Accuracy scores were drawn from the incongruent block and the mixed block. Although the window of time in which children could respond (i.e., 750 ms) was based on previous research (Davidson et al., 2006), the pacing for the mixed block was too rapid for children in this study, resulting in many missing RT scores during this block. Consequently, RT scores were drawn only from the incongruent block."

#### Multi-Source Interference Test

"The Multi-Source Interference Test (MSIT) is a measure of inhibitory control skills that is used in middle childhood and adolescence (Bush & Shin, 2006; Liu, Angstadt, Taylor, & Fitzgerald, 2016; Ursache, Noble, & Blair, 2015). There were two blocks: (1) 24 congruent trials and (2) 24 incongruent trials. On both blocks, students were presented with a sequence of three digits. For each trial, two of these digits (the distractors) were the same and one (the target) differed from the distractors (e.g., “2 2 1”). Students were instructed to press a button whose numeric value corresponded to the numeric value of the target. For example, the correct response to the sequence “2 2 1” would be “1”. For the congruent trials, the distractors were always zeroes and the numeric value of the target always corresponded to the numeric value of the correct button press (i.e., “1 0 0”, “0 2 0”, “0 0 3”). For the incongruent trials, the distractors were non-zero and the numeric value of correct button press was always different from the position of the correct response (e.g., “2 3 3”, “2 2 1”, “1 3 1”). Accuracy and RT scores from the incongruent block of trials were used."

#### Digit Span Backward

"The Digit Span Backward (DSB) is a standard measure of working memory drawn from the Wechsler Intelligence Scale for Children-IV (Flanagan & Kaufman, 2009) that is commonly used in middle childhood (Blankenship & Bell, 2015; Brocki & Bohlin, 2004; St Clair-Thompson & Gathercole, 2006). A series of digits were presented sequentially on the tablet screen. The student was instructed to enter those numbers backwards onto a numeric keypad after the last digit was presented. There were four practice trials, each using strings that were two digits long. These practice trials were followed by eight test trials of increasing difficulty (two trials each of length two, three, four, and five digits). Accuracy scores were computed as the proportion of correct test trials."

#### Scoring of EF Tasks

"Anticipatory responses—defined as a response <200 milliseconds (ms) after stimulus presentation—were recoded as missing for the accuracy scores and RT scores. Further, the H&F and MSIT tasks were timed, such that students were unable to respond after 750 ms and 2500 ms, respectively. If the student failed to respond during this window, the trial was counted as incorrect for the accuracy score and as missing for the RT score. Finally, as is standard practice, RT scores were calculated only for the accurate trials and were not calculated for the first trial in each block. To receive an accuracy score for each task, a participants were required to have non-missing scores for 8/8 DSB trials (because difficulty varied across trials), 5/12 H&F incongruent trials, 10/33 H&F mixed trials, and 15/45 MSIT incongruent trials. In addition, RT scores were based on a minimum of three accurate trials for the H&F incongruent block and the MSIT incongruent block. Three outliers, defined as accuracy or RT scores that were N4 SD above or below the sample mean, were Winsorized to the highest non-outlier value that was observed for that task.

Cronbach's alpha reliability for the RT variables was as follows: MSIT incongruent IA = 0.91, GA = 0.90; H&F incongruent block IA = 0.88, GA = 0.90. Since Cronbach's alpha systematically underestimates reliability of scale using binary indicators (Raykov, Dimitrov, & Asparouhov, 2010), we used tetrachoric correlations to correct for this bias in our computation of alpha coefficient for the binary accuracy variables. Alpha reliability for the accuracy variables was as follows: digit span IA = 0.85, GA = 0.82; MSIT incongruent block IA = 0.90, GA = 0.95; H&F incongruent block IA = 0.91, GA = 0.89; H&F mixed block IA = 0.92, GA = 0.91."

#### Academic Achievement

"State-administered standardized test scores were used to measure English/language arts and mathematics skills at two time points. In spring of 2013, scores were drawn from the California Standards Test (California Department of Education, 2016a), a test designed to match the state's academic content standards. In spring of 2015, scores were drawn from the Smarter Balance Assessment Consortium (California Department of Education, 2016b), a test designed to match the new Common Core State Standards."

#### Self-Regulated Classroom Behaviors

"Teachers reported on students' classroom behaviors relevant to academic success using the Teacher-Child Rating Scale (TCRS; Hightower, 1986). Each item on the task orientation (e.g., “well-organized”, “completes work”, “works well without adult support”) and frustration tolerance (e.g., “accepts things not going his/her way”, “copes well with failure”) scales was rated on a five-point scale, ranging from 1 = “not at all” to 5 = “very well”."

#### Covariates

"Child age in years, child gender (0 = male, 1 = female), and years of parent education were included as covariates" in all analyses. As explained in Tables 3 & 4 in the original paper, for analyses of mean differences between IA and GA scores as well as the predictive validity of students' IA and GA scores, the authors also included school fixed effects and child's ethnicity as covariates. Finally, the analysis of mean differences between IA and GA scores included "a dummy variable representing the assessment order effect" as a sixth covariate.

### Procedure	

The procedures outlined below (quoted and followed precisely from the original article) were used to collect this data set:

"Longitudinal data from four time points were used in this study. The participants described above (third- through fifth-graders) completed EF assessments in the fall of 2013 (Time 2), and teacher reports of students' classroom behavior were collected using an online questionnaire in the spring of 2014 (Time 3). Students' academic achievement data were obtained from school records for the previous academic year, 2012–13, when students attended second through fourth grades (Time 1), and for the following academic year, 2014–15, when students attended fourth through sixth grades (Time 4). Teachers were compensated with $95 for their participation and were given a tablet computer for classroom use."

#### Assessment of Executive Functions 

"Developmentally-appropriate and widely used EF tasks were adapted to be used on Android tablet computers. The tasks were designed to look fun and attractive to children by including cartoon pictures next to simplified task rules that students could easily read on their own. They were programmed to include a fixed number of practice trials and passcode-locked screens at designated intervals, to ensure that all students completed the tasks at a similar pace and were paying attention at the appropriate times to a research assistant who explained task instructions.

During GA, each child was given a tablet computer and all students in the classroom completed the EF tasks simultaneously. We developed a classroom procedure that enabled three research assistants (RAs) to administer EF assessment in a fashion similar to how teachers and their aides sometimes administer academic activities and tests. The lead RA, akin to a head teacher, stood at the front of the classroom, elicited students' attention, and explained the task rules using large posterboards that mimicked what students would later see on their tablet screens. The lead RA facilitated a group practice of each task rule by soliciting students' verbal responses. Afterward, all students had an opportunity to read the simplified task rules on their tablet screens and receive individualized feedback from the computer program, based on their performance on practice trials. The lead RA ensured that students completed practice and test trials at the same time by providing them with the passcodes to unlock different task blocks when all students were ready to proceed. Meanwhile, the other two RAs roamed the classroom like teacher's aides, helping students who needed technical assistance.

For IA, a subset of students completed the same tablet tasks outside of their classrooms in a quiet space such as a library room. A single RA explained the task rules to a child using a flipbook that contained the exact same images that were presented on posterboards in the GA context. Analogous to GA, the child was asked to verbally respond to two practice trials before proceeding to complete a practice block on a tablet computer. To account for practice effects, we administered IA in a counterbalanced fashion so that half of the students completed IA before GA and half of the students completed IA after GA."

### Analysis Plan

The four key analyses of interest, as quoted from the original paper, are as follows: 

"We performed a series of four analyses to validate GA and describe similarities and differences between IA and GA. First, we examined descriptive statistics and the correlations between IA and GA for each task. Second, we compared associations of EF composites with known demographic and educational covariates across the two assessment settings. Third, we examined whether there were mean differences between IA and GA performance composites while accounting for the ordering of the two assessments. Finally, we tested the unique contribution of EF performance in IA and GA settings for the prediction of self-regulated classroom behaviors and academic achievement."

Please see above for references to important data cleaning and exclusion rules, as well as covariates. 

### Differences from Original Study

Because this is a reproduction using the data from the original study, there were no differences in sample, setting, procedure, or analysis plan. The only differences involved the actual R code used in the analyses. 

```{r include=FALSE}
### Methods Addendum (Post Data Collection)

# You can comment this section out prior to final report with data collection.

#### Actual Sample
  # Sample size, demographics, data exclusions based on rules spelled out in analysis plan

#### Differences from pre-data collection methods plan
  # Any differences from what was described as the original plan, or “none”.
```  

## Results

### Data Preparation: High-Level Description

Based on the materials and analysis plan outlined above, key elements of data preparation include:

#### Data Exclusion and Filtering

* Recoding anticipatory responses (i.e., those <200 ms after stimulus presentation) as missing for accuracy and RT scores
* Recoding missing H&F and MSIT responses as incorrect for the accuracy score and missing for the RT score
* Excluding all inaccurate trials and the first trial in each block from RT scores 
* For participants who failed to meet the threshold of having non-missing scores for 8/8 DSB trials, 5/12 H&F incongruent trials, 10/33 H&F mixed trials, and 15/45 MSIT incongruent trials, recoding accuracy scores as missing
* For participants lacking at least three accurate trials for the H&F incongruent block and the MSIT incongruent block, recoding RT scores as missing
* Windsorizing outliers (i.e., those with accuracy or RT scores that were >4 SD above or below the sample mean) to the highest non-outlier value observed for that task

#### Additional Data Preparation: Composite Creation

The following three composite variables will need to be constructed from raw data: teacher-reported EF composite score, EF test accuracy composite score, and EF test RT composite score. 

### Creating a Hybrid 4-Variable Hightower/BRIEF Teacher-Report EF Composite Score

Inclusion criteria for composite (received directly from original authors):

Items: **HIGHTOWER**
  
  Scoring:
  
    1 = Not at all 
    2 = A little
    3 = Moderately well 
    4 = Well 
    5 = Very well
 
  Task Orientation, 
    Item #3: "Functions well even with distractions"

  Acting Out, 
    Item #2: "Fidgety, difficulty sitting still"

Items: **BRIEF** 
  
  Scoring:
    
    0 = Never
    1 = Sometimes
    2 = Often
 
  Cognitive Flexibility, 
    Item #5: "Resists or has trouble accepting a different way to 
    solve a problem with schoolwork, friends, chores, etc."	
 
  Working Memory,
    Item #25: "Has trouble with chores or tasks that have more than 
    one step"
    
```{r}
# Load relevant libraries
library(mjs)
library(dplyr)
library(tidyr)
library(ggplot2)
library(cowplot)
library(moments)

# Set up
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)

# Import data
setwd("~/Desktop/obradovic2017/writeup/data")
tcrs <- read.csv("PLUS_teacher_survey_variables_studentlevel.csv")
names(tcrs)

# Process data
# Sort by student id
tcrs <- tcrs %>%  
  arrange(stud_id) %>% 
  filter(duplicated(stud_id) == FALSE)

# Get rid of variable name issues (extra _) 
names(tcrs) <- sub("_$", "", names(tcrs))

# Drop unnecessary "child_num" column and last 10 columns (not necessary for analysis)
tcrs <- tcrs %>% 
  select(-child_num,
         -tcrs_act,
         -tcrs_task,
         -tcrs_asoc,
         -tcrs_peer,
         -tcrs_fru,
         -brief_flex,
         -brief_ec,
         -brief_wm,
         -HBQTCLO,
         -HBQTCON)

# Rename rows (from misc. list to "1, 2, 3...n")
row.names(tcrs) <- as.character(1:nrow(tcrs))

# Reverse all "acting out" Hightower items (so that higher scores indicate greater EF)
## Create character vector of all "acting out" Hightower variables
ao_vars <- grep("tcrs_ao", names(tcrs), value=TRUE)

## Create function to reverse all "acting out" Hightower variables
ao_rev <- function(x) {
  (6 - x)
}

## Apply function to vector of "acting out" Hightower variables (reverses values within existing columns)
tcrs <- tcrs %>% 
  mutate_at(.vars = ao_vars, .funs = ao_rev)

# Reverse all BRIEF items (so that higher scores indicate greater EF)
## Create character vector of all BRIEF variables
b_vars <- grep("brief", names(tcrs), value=TRUE)

## Create function to reverse BRIEF scores
b_rev <- function(x) {
  (2 - x)
} 

# Apply function to vector of BRIEF variables
tcrs <- tcrs %>% 
  mutate_at(.vars = b_vars, .funs = b_rev)

# Rename BRIEF items to make more legible 
# wm = working memory, cf = cognitive flexibility, ec = executive control
tcrs <- tcrs %>% 
  rename(brief_wm_1 = brief_1, 
         brief_wm_2 = brief_4, 
         brief_wm_3 = brief_7,
         brief_wm_4 = brief_10,
         brief_cf_1 = brief_2, 
         brief_cf_2 = brief_3, 
         brief_cf_3 = brief_6,
         brief_cf_4 = brief_9,
         brief_ec_1 = brief_5,
         brief_ec_2 = brief_8, 
         brief_ec_3 = brief_11,
         brief_ec_4 = brief_12)

# Make composite data frame called "hybrid" with four Hightower + BRIEF items of interest
hybrid_vars <- c("tcrs_ao2", "tcrs_to3", "brief_wm_2", "brief_cf_1")
hybrid <- tcrs[ , c("stud_id", hybrid_vars)]

# Add columns with each of the four items' z-scores
hybrid <- hybrid %>% 
  mutate(tcrs_ao2_z = as.numeric(scale(tcrs_ao2)),
         tcrs_to3_z = as.numeric(scale(tcrs_to3)),
         brief_wm_2_z = as.numeric(scale(brief_wm_2)),
         brief_cf_1_z = as.numeric(scale(brief_cf_1))
  )
head(hybrid)

# Remove rows with no data (2 total rows)
hybrid <- hybrid %>% 
  drop_na()

# Add column with composite scores
hybrid <- hybrid %>% 
  mutate(ef4 = as.numeric(rowMeans(hybrid[ , paste0(hybrid_vars, "_z")], na.rm=TRUE)))
head(hybrid)

# Check out rating frequencies
sapply(hybrid[ , hybrid_vars], table, useNA='always')

# Check out descriptive statistics
hybrid %>% 
  select(-stud_id) %>%
  mjs::desc(c(mean, sd, skewness, kurtosis, min, max))
  
# Check out alpha reliability (results show alpha = 0.81)
alphaRel(hybrid[ , paste0(hybrid_vars, "_z")])
```

### Creating a EF Test Accuracy Composite Score

*...coming soon! I anticipate this composite will be harder to create than the one above, since there is more data prep required (recoding missing data, etc.) to get things into shape before the composite can be calculated.*

### Creating a EF Test RT Composite Score

*...coming soon! I anticipate this composite will be harder to create than the one above, since there is more data prep required (recoding missing data, etc.) to get things into shape before the composite can be calculated.*

### Confirmatory Analysis

The analyses as specified in the analysis plan. 

The two main inferences in the original paper involve the mean differences between IA and GA performance composites and the predictive validity of EF performance in IA and GA settings for children's self-regulated classroom behaviors and academic achievement, both of which are justified using multilevel regression analyses. 

*To be filled in: Side-by-side graph with original graph is ideal here.*

### Exploratory Analyses

Any follow-up analyses desired (not required).  

## Discussion

### Summary of Replication Attempt

Open the discussion section with a paragraph summarizing the primary result from the confirmatory analysis and the assessment of whether it replicated, partially replicated, or failed to replicate the original result.  

### Commentary

Add open-ended commentary (if any) reflecting (a) insights from follow-up exploratory analysis, (b) assessment of the meaning of the replication (or not) - e.g., for a failure to replicate, are the differences between original and present study ones that definitely, plausibly, or are unlikely to have been moderators of the result, and (c) discussion of any objections or challenges raised by the current and original authors about the replication attempt.  None of these need to be long.
